---
title: "CellRanger input/output functions"
author: "Aaron Lun"
package: crio 
output: 
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Read and write CellRanger formats}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
self <- BiocStyle::Biocpkg("crio")
```

# Overview

CellRanger is a software suite developed by [10X Genomics](https://10xgenomics.com) for processing their single-cell genomics data.
`r self` implements several functions for reading some CellRanger output formats into Bioconductor objects.
This includes the feature-barcode matrix in either the Matrix Market or HDF5 formats and the HDF5 molecule information files.
It also provides functions for writing data back into CellRanger-formatted files.

To get started, run the usual Bioconductor installation instructions:

```r
install.packages("BiocManager")
BiocManager::install("crio")
```

# Feature-barcode count matrix

## Quick start 
 
The most important output from CellRanger is the feature-barcode count matrix.
Each row is a genomic feature, each column is a cell barcode, and each entry contains the number of UMIs assigned to the corresponding feature/barcode combination.
Let's demonstrate on some public data from the `r BiocStyle::Biocpkg("DropletTestFiles")` package:

```{r}
library(DropletTestFiles)
tarball <- getTestFile("tenx-3.0.0-pbmc_10k_protein_v3/1.0.0/filtered.tar.gz")
tmp <- tempfile()
dir.create(tmp)
untar(tarball, exdir=tmp)
list.files(tmp, recursive=TRUE)
```

We use the `readCounts()` function to read this into a `SingleCellExperiment` object:

```{r}
library(crio)
sce <- readCounts(file.path(tmp, "filtered_feature_bc_matrix"))
sce
class(assay(sce))
rowData(sce)
colData(sce)
```

## Read parameters

We use `configureSampleForReadCounts()` to customize the method used to read each sample.
For example:

- We specify the type of our format as `"mtx"` (i.e., Matrix Market).
  This is usually guessed from the path but it doesn't hurt to be explicit. 
- `dgCMatrix` objects are the "default" sparse matrix representation in R,
  but cannot support very large sparse matrices where the number of non-zero elements exceeds `.Machine$integer.max`. 
  So, we can instead use the `SVT_SparseMatrix` representation.
- We accelerate the file parsing by using multiple threads.

```{r}
my.sample <- configureSampleForReadCounts(
    file.path(tmp, "filtered_feature_bc_matrix"),
    type = "mtx",
    mtx.class = "SVT_SparseMatrix",
    mtx.threads = 2
)
sce <- readCounts(my.sample)
class(assay(sce))
```

Alternatively, let's consider using the HDF5 format for the feature-barcode count matrices.
The HDF5 format can be more convenient as all of the bits and pieces are stored in one file.
In addition, HDF5 is often more efficient to read from than text formats as parsing of text is not required.
We specify the `type=` here as `readCounts()`'s auto-guessing relies on the file extension,
but cached downloads from `r BiocStyle::Biocpkg("DropletTestFiles")` use a custom naming scheme that removes the original file extension.

```{r}
library(DropletTestFiles)
h5file <- getTestFile("tenx-3.1.0-5k_pbmc_protein_v3/1.0.0/filtered.h5")
sce <- readCounts(configureSampleForReadCounts(h5file, type = "hdf5"))
sce

# With HDF5, our matrix is file-backed instead of being loaded into memory.
# This reduces memory usage while manipulating the data as copies do not
# need to be created for each operation.
class(assay(sce))

# However, it is best to load the data into memory prior to any actual
# analyses, otherwise the computation will be very slow when data needs to be
# repeatedly pulled from disk. We could also do this by setting
# hdf5.realize=TRUE in the configuration for this sample.
assay(sce) <- as(assay(sce), "dgCMatrix")
```

We can also combine multiple samples into one `SingleCellExperiment`, possibly with sample-specific parameters.
Here, we only keep the intersection of row names across all samples.
We can delay the `cbind()` operation when combining matrices to reduce memory usage.

```{r}
sce <- readCounts(
    list(
        FOO = file.path(tmp, "filtered_feature_bc_matrix"),
        BAR = configureSampleForReadCounts(h5file, type = "hdf5", hdf5.realize = TRUE)
    ),
    intersect.rows = TRUE,
    delayed = TRUE
)
sce

assay(sce)
colData(sce)
```

## Writing to disk

Given a `SingleCellExperiment` object, we can write its contents to a CellRanger-compatible file format.
First, let's mock something up:

```{r}
# Mocking up some count data.
library(Matrix)
my.counts <- matrix(rpois(1000, lambda=5), ncol=10, nrow=100)
my.counts <- as(my.counts, "dgCMatrix")

library(SummarizedExperiment)
se <- SummarizedExperiment(my.counts)
colnames(se) <- paste0("BARCODE-", seq_len(ncol(my.counts)))
rownames(se) <- paste0("ENSG0000", seq_len(nrow(my.counts)))
rowData(se)$name <- paste0("GENE", seq_len(nrow(my.counts)))
rowData(se)$type <- "Gene Expression"
```

Now we write it to disk.
Here, we'll use the Matrix Market format, but we can also save to HDF5 if we want.

```{r}
tmp <- tempfile()
writeCounts(se, path=tmp, type="mtx")
list.files(tmp, recursive=TRUE)
```

Reloading it is straightforward:

```{r}
readCounts(tmp)
```

In principle, it is possible to re-use the files created by `writeCounts()` in external applications like `cellranger reanalyze`, 
but mileage may vary due to some undocumented aspects of 10X's file specifications. 
If you are running into interoperability issues, file an [issue](https://github.com/LTLA/crio) and we'll see what we can do.

# Molecule information

## Quick start

The molecule information file contains - as its name suggests - information about all of the successfully sequenced (transcript) molecules.
This is occasionally useful for troubleshooting, e.g., to diagnose issues from barcode swapping or chimeric reads. 
Let's fetch one of these files and load it in:

```{r}
molpath <- getTestFile("tenx-2.1.0-pbmc4k/1.0.0/mol_info.h5")

# Let's also load in molecules that weren't mapped.
molinfo <- readMolecules(molpath, keep.unmapped=TRUE)
molinfo$molecules

# Each entry of the 'feature' column refers to a row of 'features'.
molinfo$features
```

Now we can do some interesting things like counting the number of unassigned molecules or the average coverage of each molecule:

```{r}
# Unmapped molecules have 'feature' set to the number of features plus 1.
sum(molinfo$molecules$feature == nrow(molinfo$features) + 1L)

# 'count' contains the read coverage for each molecule. 
mean(molinfo$molecules$count)
```

The UMI sequence is stored as a 2-bit encoding, which we can decode into the base sequence if we know the UMI length.
You'll have to look this up in 10X's sequence design - for this particular dataset, I think it's 10 nt.

```{r}
decodeSequences(molinfo$molecules$umi[1:20], length=10)
```

## Making a count matrix

We can convert the molecule information `DataFrame` back into a count matrix with `countMolecules()`.
This can be used to obtain counts for further analysis after filtering out certain molecules,
e.g., those without enough read coverage or after using a more careful UMI deduplication scheme.

```{r}
counts <- countMolecules(nrow(molinfo$features), molinfo$molecules$feature, molinfo$molecules$barcode)
dim(counts)
```

In fact, we can more-or-less recover the `SingleCellExperiment` that we would have gotten from `readCounts()` on the raw count matrix (i.e., before cell calling).

```{r}
sce <- SingleCellExperiment(list(counts = counts), rowData = molinfo$features)
sce$barcode <- colnames(sce)
rownames(sce) <- rowData(sce)$id
sce
```

## Writing to disk

On some rare occasions, we may want to write the molecule information to file.
This can be done with the `writeMolecules()` function:

```{r}
# Making up some molecule information.
sim <- simulateMolecules()
tmp <- tempfile(fileext=".h5")
writeMolecules(tmp, sim)
```

Now we can read it back in as if it came from CellRanger itself:

```{r}
readMolecules(tmp)
```

Again, it is theoretically possible to re-use the files created by `writeMolecules()` in external applications, 
but mileage may vary due to some undocumented aspects of 10X's file specifications. 
File any problems in the [GitHub repository](https://github.com/LTLA/crio) and we'll see what we can do.

# Session information {-}

```{r}
sessionInfo()
```
